### 2023-05-03 submission/manuscript.Rmd
Full first draft of Discussion

Beyond the discussion of whether rarefaction is appropriate for analyzing microbiome data it is worth commenting on McMurdie and Holmes's advice to use DESeq's Variance Stabilization or edgeR's Upper Quartile Log-fold Change normalization strategies. These methods have been adopted from gene expression analysis to microbiome analysis. Gene expression analysis implicitly assumes that all samples have the same genes since. While this might work in comparing healthy and diseased tissues from a cohort of patients, it does not generalize to those patients' microbiota. Microbial populations are highly patchy in their distribution. Thus, a zero count for gene expression is more likely to represent a gene below the limit of detection whereas a zero count for a microbiome analysis is more likely to represent the true absense of the OTU. An example of where this is relevant is the necessity of adding a pseudocount to all OTUs to perform both the DESeq and edgeR-based normalizations. In WNWN, a pseudocount of 1 was used. However, this value is arbitrary and the sensitivity of the results can vary based on the patchiness of the communties being analyzed. Beyond

The choice of distance metric is a complicated question and the use of six different metrics in WNWN illustrates the challenges faced in selecting one metric. In the current analysis, the phylogenetic Weighted UniFrac distance often performed as well or better than the Bray-Cutis distances for clustering or differentiating between communities with adonis. However, the current reality is that it is computationally impractical to construct phylogenetic trees with modern datasets. Although algorithms have been developed to sidestep de novo construction of trees [XXXXXX], these algorithms depend on reference-based clustering strategies that have significant challenges [XXXXXX]. Within the ecology literature, Euclidean distances are widely avoided because joint absense is weighted the same as joint presence of taxa [XXXXXX]. Among the options analyzed here, Bray-Curtis is the most robust and practical choice. Regardless, all of the methods perform best when using rarefaction relative to the other normalization methods.

Differential abundance
* Other half of the paper
* If one OTU goes up, then others' relative abundances will all go down
* We can't help but use (rarefied) relative abundances
* Beyond the scope of this paper... recommed subsampling, performing test, and repeating to see whether results vary; in our experience the largest differences in subsamples are at low relative abundances, which are unlikely to be significant (biologically or statistically)

How we actually pick a threshold
 [**Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in acutal microbiome studies the selection of a sampling depth is not as arbitrary as the authors claim. Rather, to avoid p-hacking, researchers pick a set of criteria where they will include or exclude samples prior to testing their data.**]

Gratidue that code was published with paper
* This made it straight forward to notice that only a single subsampling step was performed for each random seed and that only 5 random seeds were used.

>>>>

Beyond the discussion of whether rarefaction is appropriate for analyzing microbiome data it is worth commenting on McMurdie and Holmes's advice to use DESeq's Variance Stabilization or edgeR's Upper Quartile Log-fold Change normalization strategies. These methods have been adopted from gene expression analysis to microbiome analysis. Gene expression analysis implicitly assumes that all samples have the same genes since. While this might work in comparing healthy and diseased tissues from a cohort of patients, it does not generalize to those patients' microbiota. Microbial populations are highly patchy in their distribution. Thus, a zero count for gene expression is more likely to represent a gene below the limit of detection whereas a zero count for a microbiome analysis is more likely to represent the true absense of the OTU. An example of where this is relevant is the necessity of adding a pseudocount to all OTUs to perform both the edgeR and DESeq-based normalizations (L443 and L487, respectively). In WNWN, a pseudocount of 1 was used. However, this value is arbitrary and the sensitivity of the results can vary based on the patchiness of the communties being analyzed. Since WNWN was published, compositional approaches have been proposed to account for uneven sampling and to provide improved interpretability [XXXXXX]. However, these methods also often require the use of pseudocounts and are not actually insensitive to uneven sampling [XXXXXXXX]. Rarefaction is preferred to these alternatives.

The choice of distance metric is a complicated question and the use of six different metrics in WNWN illustrates the challenges. Within the ecology literature, Euclidean distances are widely avoided because joint absense is weighted the same as joint presence of taxa [XXXXXX]. As discussed in point 11, metrics that are based on community membership (i.e., the presence or absense of taxa) performed worse than those than those that were based on community structure (i.e., their relative abundance). For this reason, the Unweighted UniFrac and other metrics like the Jaccard or Sorenson distance coefficients should likely be avoided. The Poisson distance metric is largely novel to the microbial ecology literature and performed no better than the more traditional metrics. In the current analysis, the phylogenetic Weighted UniFrac distance performed comparably to Bray-Cutis distances for clustering or differentiating between communities with adonis. In practice, however, the current reality is that it is computationally impractical to construct phylogenetic trees with modern datasets. Although algorithms have been developed to sidestep *de novo* construction of trees [XXXXXX], these algorithms depend on reference-based clustering strategies that have significant challenges [XXXXXX]. Among the options analyzed here, Bray-Curtis is the most robust and practical choice. Regardless, all of the methods perform best when using rarefaction relative to the other normalization methods.

Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in acutal microbiome studies the selection of a sampling depth is not as arbitrary as the authors claim. Rather, to avoid "p-hacking", researchers pick a set of criteria where they will include or exclude samples prior to testing their data. Examples of criteria might include the presence of a large gap in the sequencing effort distribution, a desire to include poorly sequenced controls, or the *a priori* stipulation of a minimum sequencing effort. To mitigate concerns of arbitrary or engineered minimum library sizes, researchers should indicate the rationale for the threshold they selected.

WNWN performed a second set of simulations to address the effect of normalization method on the ability to correctly detect differential abundance of OTUs that were randomly selected to have their relative abundances changed (i.e., Simulation B). Re-addresssing this set of simulations is beyond the scope of the current anaylsis and others have already contributed critiques [XXXXXXX]. However, many of the same concerns addressed above would apply. Perhaps more importantly is the fact that if the relative abundance of several OTUs increase, then the relative abundance of all other OTUs would necessarily decrease. Thus, one would expect every OTU to be differentially abundant. Because of this, it is not possible to truly modify the abundance of a set of OTUs indepdent of all othter OTUs. This is an important limitation of tests of methods attempting to detect differentially abundant OTUs. Regardless, a form of rarefaction could still be employed for detecting differential abundance. One could subsmaple the data, perform the statistical test, and identify the differentially abundant OTUs. This process could be repeated. In my experience the largest differences between subsamples are for low relative abundances OTUs, which are unlikely to be biologically or statistically significant.

I am grateful to McMurdie and Holmes for providing the source code that they used to conduct their analysis as R-flavored markdown files. This provided me with a better understanding of their methods including noticing that only a single subsampling step was performed for each random seed and that only 5 random seeds were used. Furthermore, their code enabled me to replicate and build upon their simulations. I have used alternative simulation and evaluation strategies to look closer at rarefaction and its alternatives and the practice of filtering low abundance seqeunces. The results of those studies are similar to this reanalysis. Rarefaction of unfiltered datasets yields the most robust results.

## Methods

### 2023-05-03 submission/manuscript.tex
Full first draft of Discussion

Problems with recommended approaches * Assume zeros are present but
below limit of detection * Differ from traditional approaches with
longstanding history in the field * Tree-based methods * Have talked
about filtering elsewhere

Differential abundance * Other half of the paper * If one OTU goes up,
then others' relative abundances will all go down * We can't help but
use (rarefied) relative abundances * Beyond the scope of this
paper\ldots{} recommed subsampling, performing test, and repeating to
see whether results vary; in our experience the largest differences in
subsamples are at low relative abundances, which are unlikely to be
significant (biologically or statistically)

\begin{itemize}
\item
  How we actually pick a threshold {[}\textbf{Although the authors claim
  that ``Rarefying counts requires an arbitrary selection of a library
  size minimum that affects downstream inference'' (page 8, column 1,
  point 3), in acutal microbiome studies the selection of a sampling
  depth is not as arbitrary as the authors claim. Rather, to avoid
  p-hacking, researchers pick a set of criteria where they will include
  or exclude samples prior to testing their data.}{]}
\item
  Gratidue that code was published with paper. This made it straight
  forward to notice that only a single subsampling step was performed
  for each random seed and that only 5 random seeds were used.
\end{itemize}

>>>>

Beyond the discussion of whether rarefaction is appropriate for
analyzing microbiome data it is worth commenting on McMurdie and
Holmes's advice to use DESeq's Variance Stabilization or edgeR's Upper
Quartile Log-fold Change normalization strategies. These methods have
been adopted from gene expression analysis to microbiome analysis. Gene
expression analysis implicitly assumes that all samples have the same
genes since. While this might work in comparing healthy and diseased
tissues from a cohort of patients, it does not generalize to those
patients' microbiota. Microbial populations are highly patchy in their
distribution. Thus, a zero count for gene expression is more likely to
represent a gene below the limit of detection whereas a zero count for a
microbiome analysis is more likely to represent the true absense of the
OTU. An example of where this is relevant is the necessity of adding a
pseudocount to all OTUs to perform both the edgeR and DESeq-based
normalizations (L443 and L487, respectively). In WNWN, a pseudocount of
1 was used. However, this value is arbitrary and the sensitivity of the
results can vary based on the patchiness of the communties being
analyzed. Since WNWN was published, compositional approaches have been
proposed to account for uneven sampling and to provide improved
interpretability {[}XXXXXX{]}. However, these methods also often require
the use of pseudocounts and are not actually insensitive to uneven
sampling {[}XXXXXXXX{]}. Rarefaction is preferred to these alternatives.

The choice of distance metric is a complicated question and the use of
six different metrics in WNWN illustrates the challenges. Within the
ecology literature, Euclidean distances are widely avoided because joint
absense is weighted the same as joint presence of taxa {[}XXXXXX{]}. As
discussed in point 11, metrics that are based on community membership
(i.e., the presence or absense of taxa) performed worse than those than
those that were based on community structure (i.e., their relative
abundance). For this reason, the Unweighted UniFrac and other metrics
like the Jaccard or Sorenson distance coefficients should likely be
avoided. The Poisson distance metric is largely novel to the microbial
ecology literature and performed no better than the more traditional
metrics. In the current analysis, the phylogenetic Weighted UniFrac
distance performed comparably to Bray-Cutis distances for clustering or
differentiating between communities with adonis. In practice, however,
the current reality is that it is computationally impractical to
construct phylogenetic trees with modern datasets. Although algorithms
have been developed to sidestep \emph{de novo} construction of trees
{[}XXXXXX{]}, these algorithms depend on reference-based clustering
strategies that have significant challenges {[}XXXXXX{]}. Among the
options analyzed here, Bray-Curtis is the most robust and practical
choice. Regardless, all of the methods perform best when using
rarefaction relative to the other normalization methods.

Although the authors claim that ``Rarefying counts requires an arbitrary
selection of a library size minimum that affects downstream inference''
(page 8, column 1, point 3), in acutal microbiome studies the selection
of a sampling depth is not as arbitrary as the authors claim. Rather, to
avoid ``p-hacking'', researchers pick a set of criteria where they will
include or exclude samples prior to testing their data. Examples of
criteria might include the presence of a large gap in the sequencing
effort distribution, a desire to include poorly sequenced controls, or
the \emph{a priori} stipulation of a minimum sequencing effort. To
mitigate concerns of arbitrary or engineered minimum library sizes,
researchers should indicate the rationale for the threshold they
selected.

WNWN performed a second set of simulations to address the effect of
normalization method on the ability to correctly detect differential
abundance of OTUs that were randomly selected to have their relative
abundances changed (i.e., Simulation B). Re-addresssing this set of
simulations is beyond the scope of the current anaylsis and others have
already contributed critiques {[}XXXXXXX{]}. However, many of the same
concerns addressed above would apply. Perhaps more importantly is the
fact that if the relative abundance of several OTUs increase, then the
relative abundance of all other OTUs would necessarily decrease. Thus,
one would expect every OTU to be differentially abundant. Because of
this, it is not possible to truly modify the abundance of a set of OTUs
indepdent of all othter OTUs. This is an important limitation of tests
of methods attempting to detect differentially abundant OTUs.
Regardless, a form of rarefaction could still be employed for detecting
differential abundance. One could subsmaple the data, perform the
statistical test, and identify the differentially abundant OTUs. This
process could be repeated. In my experience the largest differences
between subsamples are for low relative abundances OTUs, which are
unlikely to be biologically or statistically significant.

I am grateful to McMurdie and Holmes for providing the source code that
they used to conduct their analysis as R-flavored markdown files. This
provided me with a better understanding of their methods including
noticing that only a single subsampling step was performed for each
random seed and that only 5 random seeds were used. Furthermore, their
code enabled me to replicate and build upon their simulations. I have
used alternative simulation and evaluation strategies to look closer at
rarefaction and its alternatives and the practice of filtering low
abundance seqeunces. The results of those studies are similar to this
reanalysis. Rarefaction of unfiltered datasets yields the most robust
results.

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}
